---
title: "CYCLISTIC BIKES"
author: "LEANDRO ALVES"
date: "2023-12-27"
format: 
  revealjs:
    smaller: true
    scrollable: true
editor_options: 
  chunk_output_type: console
---

```{r reading_file, message=FALSE, echo=FALSE, warning=FALSE}
trip_for_analysis <- read.csv("trip_for_analysis.csv")

```

```{r Bibliotecas, message=FALSE, warning=FALSE, echo=FALSE}
library(pacman)

pacman::p_load("tidyverse", "openxlsx", "flextable", "janitor", "patchwork", "readxl", "infer", "camcorder")


```

## AGENDA

-   QUESTION DO BE ANSWERED / SOURCE
-   FINDINGS
-   CONCLUSION

## QUESTION TO BE ANSWERED

**"HOW DO MEMBER RIDERS DIFFERENTIATE FROM CASUAL USERS?"**

### SOURCE

https://divvy-tripdata.s3.amazonaws.com/index.html\
The data has been made available by Motivate International Inc.

## FINDINGS I

```{r usertype-dow-month-tile}

fg_tile_var_trip <- # for tile graph var from trip_for_analysis
trip_for_analysis %>% 
  mutate(day_of_week = wday(started_at, label = TRUE),
         mth_lb = month(started_at, label = TRUE)) %>%     
  count(day_of_week, usertype, mth_lb) 

fg_tile_var_trip %>% 
ggplot(aes(day_of_week, mth_lb, fill = n)) +
  geom_tile() +
  # geom_rect(
  #   data = fg_tile_var_trip %>% filter(usertype == "member"),
  #   aes(xmin = 1, xmax = 7, ymin = 3, ymax = 12),
  #   color = "black",
  #   fill = NA
  #   ) +
  # geom_segment(
  #       data = fg_tile_var_trip %>% filter(usertype == "member"),
  #   aes(x = 7, xend = 9, y = 6, yend = 4),
  #   color = "black",
  #   fill = NA
  # ) + 
  labs(
    title = "Difference between Members and Casual users",
    subtitle = "Regarding total of dayly use (sum of trips by day)",
    x = "",
    y = "",
    fill = "Trips"
  ) +
  facet_wrap( ~ usertype) +
  coord_equal() +
  scale_fill_viridis_c(labels = scales::label_comma(scale = 1e-3, suffix = "k")) +
  theme_minimal() +
  theme(
    # panel.background = element_rect(fill = 'transparent'),
    plot.background = element_rect(fill = 'transparent', color = NA),
    # panel.border = element_rect(fill = NA, color = NA),
    panel.grid = element_blank()
  )
```

## FINDINGS II

```{r usertype-day_of_week-hour-hist}

trip_for_analysis %>% 
  mutate(dh = hour(started_at)) %>% 
  mutate(day_of_week = wday(started_at, label = TRUE),
         mth_lb = month(started_at, label = TRUE)) %>%     
  # filter(usertype == "member") %>% 
  ggplot(aes(dh)) +
  geom_bar(color = "grey30", fill = "white") +
  labs(
    title = "Difference between Members and Casual users",
    subtitle = "Regarding total of hourly use (sum of trips by hour)",
    x = "Hours of the Day",
    y = "Number of Trips",
    # fill = "Trips"
  ) +
  facet_grid(usertype ~ day_of_week) +
  scale_y_continuous(label = scales::label_comma(scale = 1e-3, suffix = "k")) +
  theme_dark() +
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "grey30")
  )
```

## FINDINGS III

```{r usertype-ride-hist-facet-q13}

trip_for_analysis %>% 
  filter(ride_length <= quantile(trip_for_analysis$ride_length[trip_for_analysis$usertype == "member"])[4]) %>% 
  mutate(ride_length = as.integer(ride_length/60)) %>% 
ggplot(aes(ride_length, fill = usertype)) +
  geom_histogram(color = "ghostwhite", show.legend = FALSE, bins = 16) +
  labs(
    title = "Difference between Members and Casual users",
    subtitle = "Regarding to ride length for first 75% of total users",
    y = "Number of Trips",
    x = "ride_length(min)"
  ) +
  # scale_x_log10() +
  facet_wrap( ~ usertype, scales = "free_y") +
  scale_y_continuous(label = scales::label_comma(scale = 1e-3, suffix = "k")) +
  scale_fill_manual(values = c("skyblue4", "grey70")) +
  theme_minimal()
```

## CONCLUSION {.smaller}

Casual users do longer trips and mostly on weekends. By the other hand, member users do shorter trips mainly on business days and mainly for work.

The **typical user** is Male (`r round(mean(trip_for_analysis$gender[trip_for_analysis$usertype == "member"] == "Male", na.rm = TRUE)*100, 1)`%) and his Average Age is `r round(mean(trip_for_analysis$age[trip_for_analysis$usertype == "member"]), 1)` years.

-   Regarding the **ride length** it was found the following differences:\
    Member Users:

    -   Typically under `r round(quantile(trip_for_analysis$ride_length[trip_for_analysis$usertype == "member"])[4]/60, 1)` minutes (75%).

    -   Average: `r round(mean(trip_for_analysis$ride_length[trip_for_analysis$usertype == "member"])/60, 1)` minutes.

    -   Median: `r round(quantile(trip_for_analysis$ride_length[trip_for_analysis$usertype == "member"])[3]/60, 1)` minutes.

-   Casual Users:

    -   Typically under `r round(quantile(trip_for_analysis$ride_length[trip_for_analysis$usertype == "casual"])[4]/60, 1)` minutes (75%).

    -   \- Average: `r round(mean(trip_for_analysis$ride_length[trip_for_analysis$usertype == "casual"])/60, 1)` minutes.

    -   \- Median: `r round(quantile(trip_for_analysis$ride_length[trip_for_analysis$usertype == "casual"])[3]/60, 1)` minutes.

## SUGESTION FOR FURTHER ACTION

-   to identify casual users that fits the profile and communicate them some ofert.\
-   to identify the typical stations the members starts from more often and do some marketing appeal within.\
-   to make some generic campaings on using bikes for work as a good idea complying with ESG innitiatives.
